A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods

### 2024-06-06 Draft/Draft_MEM.tex
Updating draft.

\abstract{
\begin{quote}
Monitoring of influenza activity is a key task of public health agencies around the world. Intensity thresholds are a commonly used tool to retrospectively classify season peak intensity and to compare current influenza activity to the peak intensity of past seasons. The resulting classifications into low, medium, high or very high intensity help to inform national risk assessment, public health planning and ressource allocation. Moreover, they feed into global risk analyses like the weekly World Health Organization (WHO) Influenza Update. Two common approaches to compute intensity thresholds, recommended in dedicated WHO guidelines, are the  % A common approach to assess season peak activity qualitatively is via intensity thresholds, which are derived empirically from peak intensities of recent seasons. to make influenza activity comparable across countries and surveillance systems. 
moving epidemic method (MEM) and the WHO method. In both methods, thresholds correspond to quantiles of a normal distribution which has been fitted to a set of potentially transformed historical observations. While an extensive body of literature on practical applications of the MEM and WHO method exists, their statistical properties have not been assessed systematically. In this paper we study the two methods analytically and in a simulation study based on re-sampling of data from French and US routine influenza surveillance. Moreover, straightforward extensions to account for small sample sizes and secular trends are described. We find that under the default settings, both the MEM and WHO method on average classify more season peaks than intended as high or very high intensity. Combining characteristics of both and adding a standard correction for small sample sizes, better-calibrated thresholds can be achieved. Even these, however, have modest sensitivity and positive predictive values. This concerns especially thresholds for the very high intensity level, which should thus be interpreted with care.
\end{quote}
}

@@

Intensity thresholds form part of monitoring procedures in numerous countries, contributing to national-level risk appraisal and planning; see e.g., \cite{CDC2024} on their use in the United States. In addition, intensity classifications feed into international situation assessments by organizations like WHO or the European Center for Disease Prevention and Control (ECDC). In this context, they facilitate comparisons and summaries across countries and surveillance systems. In practice, two uses of thresholds can be distinguished \citep{CDC2024}:
\begin{itemize}
\item[(1)] In the aftermath of an influenza season, thresholds can be used to classify its peak intensity. As an example, the top panel of Figure \ref{fig:maps} shows a retrospective assessment of the 2022/23 season by the US Centers for Disease Prevention and Control (CDC; \citealt{White2023}). In terms of influenza-associated outpatient visits, hospitalization rates, and deaths among children and adolescents, this season was judged as high intensity.
\item[(2)] While an influenza season is ongoing, values of relevant indicators can be compared to intensity thresholds on a weekly basis. We note that the resulting classification is relative to past peaks rather than typical values at a given time of the season. This use of thresholds is illustrated in the bottom panel of Figure \ref{fig:maps}. In this display, published by WHO and ECDC on the \textit{FluNewsEurope} website (\url{https://flunewseurope.org/}), influenza activity across Europe is summarized in a map and a heat chart.
\end{itemize}
%As will be detailed in Section \ref{sec:definitions}, the definition and desired statistical properties of thresholds refer exclusively to season peak values. Retrospective classification of seasons is thus well-aligned with their original purpose. The use in real time, on the other hand, can be seen as a pragmatic monitoring tool, and it needs to be kept in mind that the classification is relative to past peak activity rather than usual activity at a given time of the year.
% While thresholds are designed to assess season peak intensities, they are also used to monitor spatial and temporal patterns over the course of a season. To this end, intensity classifications are commonly summarized in heat charts and maps; see Figure \ref{fig:maps} for an example from the \textit{Flu News Europe} platform run by the World Health Organization (WHO) and the European Center for Disease Prevention and Control (ECDC). Intensity thresholds thus play an important role in assessing the influenza situation at an international level.

@@

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the so-called \textit{WHO method} \citep{WHO2014} and the \textit{moving epidemic method} (MEM; \citealt{Vega2015}). The latter has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied in detail. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods in their current form and suggest some extensions based on related disease monitoring methods, notably to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the recommended default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, number of season peaks classified as high or very high intensity is higher than intended. The best behaviour of thresholds is achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even the recommended configuration, however, is characterized by rather modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns in particular thresholds for very high intensity, which should thus be interpreted with care. % and should be kept in mind when using both the MEM and WHO method.

@@

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the right fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the right fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such. Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold. This aspect can only be assessed in theory and simulation studies, though; in real-world applications, there is no ``gold standard'' to determine the true intensity level of a season peak. % As a complementary perspective to sensitivity and specificity, confusion matrices can be considered to assess how peaks which are truly very high, high, medium or low intensity are classified. % The true distribution of season peaks is of course unknown in practice, but the sensitivity and specificity of a given method can be studied theoretically and in simulation studies.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method should not require advanced statistical training. The method should be simple enough that a quantitatively literate user can develop a good intuition of its functioning and parameters.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

@@

\noindent The MEM and WHO methods are special cases of this procedure, see also overview in Table \ref{tab:differences} and the graphical illustration in Figure \ref{fig:illustration}. In the MEM method, no smoothing is applied and the default data transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number of historical observations is thus kept approximately fixed. This description corresponds to that by \cite{Vega2015} and is also reflected in the default settings of the R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$ (\texttt{i.type.intensity	= 5} for no transformation, \texttt{i.type.intensity = 6} for the log transformation). We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, alternatives like LOESS are available (locally estimated scatterplot smoothing; used e.g., by \citealt{Wang2023}). % The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above. We here use it in a more narrow sense for the specification from \cite{Vega2015}, reflected in the default settings of the \texttt{mem} package.
% \begin{itemize}
% \item

The implementation of the WHO method in the publicly available \textit{WHO Average Curves} Shiny Web App \citep{WHO2023} likewise offers a lot of flexibility. Our description is based on the default settings as of May 2023 and the description in \cite{WHO2014}. Smoothing of data prior to the computation of thresholds is recommended, with a default of $l = 3$ (adapted from $l = 4$ in \citealt[p68]{WHO2014}). Subsequently, $n = 1$ observation per season is used and by default no transformation is applied to the reference set. If peak incidences vary strongly across seasons, a log transformation is recommended. At least $m = 3$ historical seasons are required, but it is noted that the ``accuracy of these thresholds should be expected to increase with the number of seasons of good quality data available'' \cite[p22]{WHO2023}. Thresholds are by default applied to unsmoothed new peak values.

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{figure/illustration_mem_who.pdf}
\end{center}
\caption{Illustration of the MEM and WHO methods with their respective default settings, computed using $m = 6$ historical seasons (solid lines) and applied to a seventh (dotted lines). Here, both methods classify the new season peak as medium intensity. Observations included in the reference set are highlighted by dots (with $n = 30/6 = 5$ for MEM, $n = 1$ smoothed values for WHO). The distribution fitted to the reference set is shown as a vertical density plot. Note that for the MEM this is a log-normal as the normal distribution is fitted to log-transformed values. The data used for illustration correspond to the years 2012--2019 from the French region of Grand Est, see Section \ref{subsec:data} and Figure \ref{fig:data}.}
\label{fig:illustration}
\end{figure}

@@

\begin{table}[h]
\caption{Default settings of the moving epidemic and WHO methods.}
\label{tab:differences}
\begin{center}
\footnotesize
\begin{tabular}{lll}
\toprule
& moving epidemic method & WHO Method \\
\midrule
smoothing of historical data & none & moving average, $l = 3$\\
number $n$ of observations used per season & $n = \max[\text{round}(30/m), 1]$ & $n = 1$\\
default transformation for reference set & natural logarithm & none\\
recommended number $m$ of historical seasons & $5 \leq m \leq 10$ & $m \geq 3$, more recommended\\
smoothing of new season peak & none & none\\
\bottomrule
\end{tabular}
\end{center}

@@

@@

The EARS approach can also be seen as an intercept-only linear regression model
$$
Y_{t} = \beta_0 + \varepsilon_{t}, \ \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
$$
with $\bar{y}$ and $s^2$ as estimators of $\beta_0$ and $\sigma_\varepsilon^2$. Indeed, regression-based approaches represent a flexible and general monitoring framework, compare Section 2 in \cite{Unkel2012}. Classic parametric approaches like the \textit{Serfling method} \citep{Serfling1896} are widely used, see e.g., \cite{Thompson2009} for an application to influenza mortality. Here, a regression model including time trends and sinusoidal functions for seasonality, say,
\begin{equation}
Y_{t} = \beta_0 + \beta_1 t + \beta_3 \sin(2\pi t / \omega) + \beta_4 \cos(2\pi t / \omega) + \varepsilon_t, \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
\label{eq:serfling}
\end{equation}
is fitted to historical values $y_1, \dots, y_m$ spanning multiple seasons. The period of seasonal cycles is given by $\omega$, with $\omega = 52$ for weekly data and yearly seasonality. The model fit subsequently serves to generate a prediction interval for a new value $y_{m + 1}$. The equally popular \textit{Farrington method} is conceptually similar, but adapts to low-count settings via suitable data transformations \citep{Farrington1996} or count-valued response distributions \citep{Noufaily2013}. While technically a generalization of the EARS approach, the rationale of the Serfling and Farrington methods is somewhat different. As already noted by \cite{Hutwagner2003}, the latter aim to detect deviations from past seasonal patterns while accounting for long-term trends. EARS, on the other hand, targets abrupt changes relative to recent incidence levels.

@@

Implicitly, both the WHO and the moving epidemic method assume that the elements of $\mathcal{Y}$ are independent and identically distibuted draws from a Gaussian distribution, and that a new peak value $Y_{m + 1}^{(1)}$ comes from that same distribution. As we will examine different violations of this overall assumption, we state its various aspects in more detail. To this end denote by
\begin{equation}
\tilde{\mathbf{Y}}_j = (Y^{(1)}_j, \dots, Y^{(n)}_j)^\top\label{eq:Y_tilde}
\end{equation}
the random vector of the $n$ largest transformed incidence values from season $j$ in decreasing order. The set $\mathcal{Y}$ thus results from pooling $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_m$. We then distinguish the following assumptions.
\begin{itemize}
\item[\textbf{(H)}] It is assumed that the elements of $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_{m + 1}$ are \textit{homogeneous} in two ways.
\begin{itemize}
\item[\textbf{(HB)}] \textit{Between-season} homogeneity is assumed, meaning that the distribution of $\tilde{\mathbf{Y}}_j$ is the same for all $j = 1, \dots, m + 1$. Secular time trends and other changes in seasonal patterns are thus excluded.
\item[\textbf{(HW)}] Moreover, \textit{within-season} homogeneity is assumed in the sense that the elements $Y_j^{(1)}, \dots, Y_j^{(n)}$ of each vector $\tilde{\mathbf{Y}}_j, j = 1, \dots, m$ all follow the same distribution.
\end{itemize}
\item[\textbf{(I)}] Two \textit{independence} assumptions paralleling (HB) and (HW) are made.
\begin{itemize}
\item[\textbf{(IB)}] The vectors $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_{m + 1}$ are assumed to be independent. We thus do not allow for for correlations between subsequent seasons.
\item[\textbf{(IW)}] The elements $Y_j^{(1)}, \dots, Y_j^{(n)}$ within a vector $\tilde{\mathbf{Y}}_j$, i.e., the $n$ highest observations from the same season $j$, are assumed to be independent.
\end{itemize}
\item[\textbf{(N)}] It is assumed that the transformation $f$ leads to multivariate \textit{normality} of the vectors $\tilde{\mathbf{Y}}_j, j = 1, \dots, m$.
\end{itemize}
Whether assumptions (HB) and (IB) are reasonable is largely an empirical question. Assumptions (HW) and (IW), on the other hand, are almost inevitably violated if $n > 1$, as by construction $Y_j^{(1)} \geq Y_j^{(1)} \geq \dots \geq Y_j^{(n)}$. Technically, assumption (N) is likewise incompatible with this constraint, but may be helpful in practice.  For the time being we nonetheless accept all of the above assumptions and state standard theory on predictive quantiles in this setting. For large $m\times n$, the normal approximation
$$
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd
$$
as in equation \eqref{eq:def_q} can be used. A common rule of thumb is to require $m \times n \geq 30$ (which seems to be the reasoning behind the MEM default setting $n = 30/m$). For small $m \times n$, however,
\begin{equation*}
\hat{q}_{Y, \alpha} = \bar{y} + t_{m\times n - 1, \alpha} \times \sqrt{1 + \frac{1}{m\times n}} \times s
\end{equation*}
as in equation \eqref{eq:q_Y_t} should be used (see e.g., \citealt{Preston2000}). The $t$-distribution with $m \times n$ degrees of freedom here accounts for the fact that the standard deviation $s$ needs to be estimated along with $\bar{y}$. As illustrated in Figure \ref{fig:illustration_t}, the predictive quantiles, and thus intensity thresholds, resulting from \eqref{eq:def_q} and \eqref{eq:q_Y_t} can differ substantially, with the latter leading to higher values for $\alpha = 0.9, 0.975$. We note that \cite{Allevius2020} have brought forward the same argument with respect to the EARS method.

@@

As mentioned in Secion \ref{sec:related_literature}, regression approaches are a natural extension of moment-based techniques like the WHO and moving epidemic method. We illustrate this with a suggestion on how to account for secular trends in threshold setting, thus relaxing assumption (HB). Secular trends are the main reasons why for the MEM it is recommend to use no more than $m = 10$ seasons of historical data, with \citeauthor{Vega2013} (\citeyear{Vega2013}, p.556) cautioning that ``more than ten seasons may further increase accuracy but make a model susceptible to biases from secular trends.'' An obvious model accouning for a secular trend is
\begin{equation}
Y_j^{(1)} = \beta_0 + \beta_1 \times j + \epsilon_j, \ \ \ \ \varepsilon_j \stackrel{\text{i.i.d.}}{\sim} \text{N}(0, \sigma^2_\varepsilon).\label{eq:trend}
\end{equation}
Parameter estimates $\hat{\alpha}$ and $\hat{\beta}$ can be computed using peak values $Y_1^{(1)}, \dots, Y_m^{(1)}$ from the last $m$ seasons and the ordiary least squares method. An unbiased estimator of ${\sigma}^2_\varepsilon$ is
$$
\hat{\sigma}^2_\varepsilon = \frac{1}{m - 2} \sum_{j = 1}^m (y_i - \beta_0 - \beta_1 \times j)^2.
$$
Predictive quantiles at levels $\alpha = 0.4, 0.9, 0.975$, and thus intensity thresholds, can be obtained using standard methods implemented in statistical software packages such as R. As we show in Supplement \ref{suppl:regression}, the generalization of equation \eqref{eq:q_Y_t} resulting for model \eqref{eq:trend} is of a rather simple form. A graphical illustration of the resulting time-varying thresholds can be found in Figure \ref{fig:illustration_trend}.

@@

>>>>

\abstract{
\begin{quote}
Monitoring of influenza activity is a key task of public health agencies around the world. Intensity thresholds are a commonly used tool to retrospectively classify season peak intensity and to compare current influenza activity to the peak intensity of past seasons. The resulting classifications into low, medium, high or very high intensity help to inform national risk assessment, public health planning and ressource allocation. Moreover, they feed into global risk analyses like the weekly World Health Organization (WHO) Influenza Update. Two common approaches to compute intensity thresholds, recommended in dedicated WHO guidelines, are the
moving epidemic method (MEM) and the WHO method. In both methods, thresholds correspond to quantiles of a normal distribution which has been fitted to a set of potentially transformed historical observations. While an extensive body of literature on practical applications of the MEM and WHO method exists, their statistical properties have not been assessed systematically. In this paper we study the two methods analytically and in a simulation study based on re-sampling of data from French and US routine influenza surveillance. Moreover, straightforward extensions to account for small sample sizes and secular trends are described. We find that under the default settings, both the MEM and WHO method on average classify more season peaks than intended as high or very high intensity. Combining characteristics of both and adding a standard correction for small sample sizes, better-calibrated thresholds can be achieved. Even these, however, have modest sensitivity and positive predictive values. This concerns especially thresholds for the very high intensity level, which should thus be interpreted with care.
\end{quote}
}

@@

Intensity thresholds form part of monitoring procedures in numerous countries, contributing to national-level risk appraisal and planning; see e.g., \cite{CDC2024} on their use in the United States. In addition, intensity classifications feed into international situation assessments by international organizations like the weekly \textit{WHO Influenza Update} (see e.g., \citealt{WHO2024}). In this context, they facilitate comparisons and summaries across countries and surveillance systems. In practice, two uses of thresholds can be distinguished \citep{CDC2024}:
\begin{itemize}
\item[(1)] In the aftermath of an influenza season, thresholds can be used to classify its peak intensity. As an example, the top panel of Figure \ref{fig:maps} shows a retrospective assessment of the 2022/23 season by the US Centers for Disease Prevention and Control (CDC; \citealt{White2023}). In terms of influenza-associated outpatient visits, hospitalization rates, and deaths among children and adolescents, this season was judged as high intensity.
\item[(2)] While an influenza season is ongoing, values of relevant indicators can be compared to intensity thresholds on a weekly basis. We note that the resulting classification is relative to past peaks rather than typical values at a given time of the season. This use of thresholds is illustrated in the bottom panel of Figure \ref{fig:maps}. In this display, published by WHO and ECDC on the \textit{FluNewsEurope} website (\url{https://flunewseurope.org/}), influenza activity across Europe is summarized in a map and a heat chart.
\end{itemize}
%As will be detailed in Section \ref{sec:definitions}, the definition and desired statistical properties of thresholds refer exclusively to season peak values. Retrospective classification of seasons is thus well-aligned with their original purpose. The use in real time, on the other hand, can be seen as a pragmatic monitoring tool, and it needs to be kept in mind that the classification is relative to past peak activity rather than usual activity at a given time of the year.
% While thresholds are designed to assess season peak intensities, they are also used to monitor spatial and temporal patterns over the course of a season. To this end, intensity classifications are commonly summarized in heat charts and maps; see Figure \ref{fig:maps} for an example from the \textit{Flu News Europe} platform run by the World Health Organization (WHO) and the European Center for Disease Prevention and Control (ECDC). Intensity thresholds thus play an important role in assessing the influenza situation at an international level.

@@

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the so-called \textit{WHO method} \citep{WHO2014} and the \textit{moving epidemic method} (MEM; \citealt{Vega2015}). The latter has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied in detail. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods in their current form and suggest some extensions based on related disease monitoring methods, notably to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the recommended default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, the number of season peaks classified as high or very high intensity is higher than intended. The best behaviour of thresholds is achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even the recommended configuration, however, is characterized by rather modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns in particular thresholds for very high intensity, which should thus be interpreted with care. % and should be kept in mind when using both the MEM and WHO method.

@@

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the right fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the right fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such. Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold. This aspect, though can only be assessed in theory and simulation studies, where the true intensity status can be determined based on the assumed distribution of season peaks. In real-world applications, there is no ``gold standard'' to determine the true intensity level. % As a complementary perspective to sensitivity and specificity, confusion matrices can be considered to assess how peaks which are truly very high, high, medium or low intensity are classified. % The true distribution of season peaks is of course unknown in practice, but the sensitivity and specificity of a given method can be studied theoretically and in simulation studies.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method should not require advanced statistical training. The method should be simple enough that a quantitatively literate user can develop a good intuition of its functioning and parameters.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

@@

\noindent The MEM and WHO methods are special cases of this procedure, see also overview in Table \ref{tab:differences} and the graphical illustration in Figure \ref{fig:illustration}. In the MEM method, no smoothing is applied and the default data transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number of historical observations is thus kept approximately fixed. This description corresponds to that by \cite{Vega2015} and is also reflected in the default settings of the R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$, see Appendix \ref{suppl:mapping}. We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, alternatives like LOESS are available (locally estimated scatterplot smoothing; used e.g., by \citealt{Wang2023}). Thresholds based on a $t$-distribution as in equation \eqref{eq:q_Y_t} were not implemented when the computations for this paper were run, and an independent custom function was used instead. In the meantime, however, this feature has been added to the \texttt{mem} package, see Appendix \ref{suppl:mapping}. % The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above. We here use it in a more narrow sense for the specification from \cite{Vega2015}, reflected in the default settings of the \texttt{mem} package.
% \begin{itemize}
% \item

The implementation of the WHO method in the publicly available \textit{WHO Average Curves} Shiny Web App \citep{WHO2023} likewise offers a lot of flexibility. Our description is based on the default settings as of June 2024 and the description in \cite{WHO2014}. Smoothing of data prior to the computation of thresholds is recommended, with a default of $l = 3$ (adapted from $l = 4$ in \citealt[p68]{WHO2014}). Subsequently, $n = 1$ observation per season is used and by default no transformation is applied to the reference set. If peak incidences vary strongly across seasons, a log transformation is recommended. At least $m = 3$ historical seasons are required, but it is noted that the ``accuracy of these thresholds should be expected to increase with the number of seasons of good quality data available'' \cite[p22]{WHO2023}. Thresholds are by default applied to unsmoothed new peak values. Thresholds based on a $t$-distribution are available, but as of June 2024 do not seem to take into account the factor $\sqrt{1 + 1/(m + n)}$ from equation \eqref{eq:q_Y_t}.

@@

\begin{table}[h]
\caption{Default settings of the moving epidemic and WHO methods. Both procedures are visualized in Figure \ref{fig:illustration}.}
\label{tab:differences}
\begin{center}
\footnotesize
\begin{tabular}{lll}
\toprule
& moving epidemic method & WHO Method \\
\midrule
smoothing of historical data & none & moving average, $l = 3$\\
number $n$ of observations used per season & $n = \max[\text{round}(30/m), 1]$ & $n = 1$\\
default transformation for reference set & natural logarithm & none\\
recommended number $m$ of historical seasons & $5 \leq m \leq 10$ & $m \geq 3$, more recommended\\
smoothing of new season peak & none & none\\
\bottomrule
\end{tabular}
\end{center}

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.78\textwidth]{figure/illustration_mem_who.pdf}
\end{center}
\caption{Illustration of the MEM and WHO methods with their respective default settings, computed using $m = 6$ historical seasons (solid lines) and applied to a seventh (dotted lines). Here, both methods classify the new season peak as medium intensity. Observations included in the reference set are highlighted by dots (with $n = 30/6 = 5$ for MEM, $n = 1$ smoothed values for WHO). The distribution fitted to the reference set is shown as a vertical density plot. Note that for the MEM this is a log-normal as the normal distribution is fitted to log-transformed values. The data used for illustration correspond to the years 2012--2019 from the French region of Grand Est, see Section \ref{subsec:data} and Figure \ref{fig:data}.}
\label{fig:illustration}
\end{figure}

@@

The EARS approach can also be seen as an intercept-only linear regression model
$$
Y_{t} = \beta_0 + \varepsilon_{t}, \ \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
$$
with $\bar{y}$ and $s^2$ as estimators of $\beta_0$ and $\sigma_\varepsilon^2$. Indeed, regression-based approaches represent a flexible and general monitoring framework, compare Section 2 in \cite{Unkel2012}. Classic parametric approaches like the \textit{Serfling method} \citep{Serfling1896} are widely used, see e.g., \cite{Thompson2009} for an application to influenza mortality. Here, a regression model including time trends and sinusoidal functions for seasonality, say,
\begin{equation}
Y_{t} = \beta_0 + \beta_1 t + \beta_3 \sin(2\pi t / \omega) + \beta_4 \cos(2\pi t / \omega) + \varepsilon_t, \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
\label{eq:serfling}
\end{equation}
is fitted to historical values $y_1, \dots, y_m$ spanning multiple seasons. The period of seasonal cycles is given by $\omega$, with $\omega = 52$ for weekly data and yearly seasonality. The model fit subsequently serves to generate a prediction interval for a new value $y_{m + 1}$. The equally popular \textit{Farrington method} is conceptually similar, but adapts to low-count settings via suitable data transformations \citep{Farrington1996} or count-valued response distributions \citep{Noufaily2013}. While technically a generalization of the EARS approach, the rationale of the Serfling and Farrington methods is somewhat different. As noted by \cite{Hutwagner2003}, the latter aim to detect deviations from past seasonal patterns while accounting for long-term trends. EARS, on the other hand, targets abrupt changes relative to the most recent observations.

@@

Implicitly, both the WHO and the moving epidemic method assume that the elements of $\mathcal{Y}$ are independent and identically distibuted draws from a Gaussian distribution, and that a new peak value $Y_{m + 1}^{(1)}$ comes from that same distribution. As we will examine different violations of this overall assumption, we state its various aspects in more detail. To this end denote by
\begin{equation}
\tilde{\mathbf{Y}}_j = (Y^{(1)}_j, \dots, Y^{(n)}_j)^\top\label{eq:Y_tilde}
\end{equation}
the random vector of the $n$ largest transformed incidence values from season $j$ in decreasing order. The set $\mathcal{Y}$ thus results from pooling $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_m$. We then distinguish the following assumptions.
\begin{itemize}
\item[\textbf{(H)}] It is assumed that the elements of $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_{m + 1}$ are \textit{homogeneous} in two ways.
\begin{itemize}
\item[\textbf{(HB)}] \textit{Between-season} homogeneity is assumed, meaning that the distribution of $\tilde{\mathbf{Y}}_j$ is the same for all $j = 1, \dots, m + 1$. Secular time trends and other changes in seasonal patterns are thus excluded.
\item[\textbf{(HW)}] Moreover, \textit{within-season} homogeneity is assumed in the sense that the elements $Y_j^{(1)}, \dots, Y_j^{(n)}$ of each vector $\tilde{\mathbf{Y}}_j, j = 1, \dots, m$ all follow the same distribution.
\end{itemize}
\item[\textbf{(I)}] Two \textit{independence} assumptions paralleling (HB) and (HW) are made.
\begin{itemize}
\item[\textbf{(IB)}] The vectors $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_{m + 1}$ are assumed to be independent. We thus do not allow for for correlations between subsequent seasons.
\item[\textbf{(IW)}] The elements $Y_j^{(1)}, \dots, Y_j^{(n)}$ within a vector $\tilde{\mathbf{Y}}_j$, i.e., the $n$ highest observations from the same season $j$, are assumed to be independent.
\end{itemize}
\item[\textbf{(N)}] It is assumed that the transformation $f$ leads to multivariate \textit{normality} of the vectors $\tilde{\mathbf{Y}}_j, j = 1, \dots, m$.
\end{itemize}
Whether assumptions (HB) and (IB) are reasonable is largely an empirical question. Assumptions (HW) and (IW), on the other hand, are almost inevitably violated if $n > 1$, as by construction $Y_j^{(1)} \geq Y_j^{(1)} \geq \dots \geq Y_j^{(n)}$. Technically, assumption (N) is likewise incompatible with this constraint, but may be helpful in practice.  For the time being we nonetheless accept all of the above assumptions and state standard theory on predictive quantiles in this setting. For large $m\times n$, the normal approximation
$$
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd
$$
as in equation \eqref{eq:def_q} can be used. A common rule of thumb is to require $m \times n \geq 30$ (which seems to be the reasoning behind the MEM default setting $n = 30/m$). For small $m \times n$, however,
\begin{equation*}
\hat{q}_{Y, \alpha} = \bar{y} + t_{m\times n - 1, \alpha} \times \sqrt{1 + \frac{1}{m\times n}} \times s
\end{equation*}
as in equation \eqref{eq:q_Y_t} should be used (see e.g., \citealt{Preston2000}). The $t$-distribution with $m \times n$ degrees of freedom here accounts for the fact that the standard deviation $s$ needs to be estimated along with $\bar{y}$. As illustrated in Figure \ref{fig:illustration_t}, the predictive quantiles, and thus intensity thresholds, resulting from \eqref{eq:def_q} and \eqref{eq:q_Y_t} can differ substantially for small $m \times n$, with the latter leading to higher values for $\alpha = 0.9, 0.975$. We note that \cite{Allevius2020} have brought forward the same argument with respect to the EARS method.

@@

As mentioned in Secion \ref{sec:related_literature}, regression approaches are a natural extension of moment-based techniques like the WHO and moving epidemic method. We illustrate this with a suggestion on how to account for secular trends in threshold setting, thus relaxing assumption (HB). Secular trends are the main reasons why for the MEM it is recommend to use no more than $m = 10$ seasons of historical data, with \citeauthor{Vega2013} (\citeyear{Vega2013}, p.556) cautioning that ``more than ten seasons may further increase accuracy but make a model susceptible to biases from secular trends.'' An obvious model accouning for a secular trend is
\begin{equation}
Y_j^{(1)} = \beta_0 + \beta_1 \times j + \varepsilon_j, \ \ \ \ \varepsilon_j \stackrel{\text{i.i.d.}}{\sim} \text{N}(0, \sigma^2_\varepsilon).\label{eq:trend}
\end{equation}
Parameter estimates $\hat{\alpha}$ and $\hat{\beta}$ can be computed using peak values $Y_1^{(1)}, \dots, Y_m^{(1)}$ from the last $m$ seasons and the ordinary least squares method. An unbiased estimator of ${\sigma}^2_\varepsilon$ is
$$
\hat{\sigma}^2_\varepsilon = \frac{1}{m - 2} \sum_{j = 1}^m (y_i - \beta_0 - \beta_1 \times j)^2.
$$
Predictive quantiles at levels $\alpha = 0.4, 0.9, 0.975$, and thus intensity thresholds, can be obtained using standard methods implemented in statistical software packages such as R. As we show in Supplement \ref{suppl:regression}, the generalization of equation \eqref{eq:q_Y_t} resulting for model \eqref{eq:trend} is of a rather simple form. A graphical illustration of the resulting time-varying thresholds can be found in Figure \ref{fig:illustration_trend}. This approach could easily be extended to accommodate $n > 1$ values from each historical season. However, as we will argue in Section \ref{subsec:choice_n}, setting $n > 1$ will lead to biased thresholds, unless a more flexible model formulation accounting for relevant correlation structures is adopted. We therefore only consider $n = 1$.

@@

\subsection{Mapping  between notation of this paper and arguments in the R package \texttt{mem}}
\label{suppl:mapping}
